---
title: "REVIVE: patient characteristics harmonization"
format:
  html:
    toc: true
execute:
  echo: true
  cache: false
---

## Purpose

Harmonize `data/Patients_characterisitics.csv` using the same structure as
`u01l_harmonization.R`: read the source, apply a dataset-specific function
with explicit variable conversions, and save a derived dataset. The source
filename above matches the spelling of the existing export.

The conversions are defined in `R/REVIVE_harmonization.R`, using
`mutate()` and `case_when()` with each source variable named explicitly.
Original columns and rows are retained alongside the harmonized variables.
The source CSV is never overwritten.

## Setup

This document uses `dplyr`, `readr`, and `knitr`. Quarto runs its R chunks
from this document's directory (`R/`).

```{r}
#| label: setup
library(dplyr)
library(readr)

source("REVIVE_harmonization.R")
```

## Read patient characteristics

Read IDs as text to preserve leading zeros. Source fields are read as text,
and numeric conversion is applied explicitly in the harmonization function.

```{r}
#| label: read-patients
patients <- read_csv(
  "../data/Patients_characterisitics.csv",
  col_types = cols(.default = col_character())
)
```

## Apply dictionary conversions

Mappings follow the local
[REDCap dictionary](../ref/Revive_dictionary.pdf), dated October 7, 2026.

- `age_calc_consent` becomes numeric `age` (age at consent).
- `group_consent_select` becomes `study_group`: 1 = Standard Cohort,
  2 = Comprehensive Cohort (field 24, page 5).
- `sex` becomes `sex_at_birth`: 1 = Male, 2 = Female,
  97 = Prefer not to answer (field 61, page 9).
- `ethnicity` becomes `hispanic_latino`: 1 = Yes, 2 = No,
  97 = Prefer not to answer (field 66, page 10).
- Race checkboxes become separate Yes/No indicators for White, Black or
  African American, Asian, American Indian or Alaska Native, Native Hawaiian
  or Other Pacific Islander, Other, and Prefer not to answer
  (field 64, page 10).
- `tx_status_yn` becomes `transplant_recipient`: 1 = Yes, 0 = No.
- Recent transplant organ checkboxes become separate Yes/No indicators:
  1 Kidney, 2 Liver, 3 Lung, 4 Heart, 5 Pancreas, 6 Intestine,
  7 Islet cell, and 88 Other (field 73, page 11).
- Each immunosuppressive medication checkbox becomes its own Yes/No variable,
  using the choices in field 92 on page 13. Multiple selections are retained
  for race, transplant organs, and medications.
- Missing or unrecognized checkbox codes remain `NA`, rather than becoming No.
- Other-organ and other-medication descriptions are retained, with surrounding
  whitespace removed and blank text represented as `NA`.

The new dictionary supplies the `recent_tx_organ` choices explicitly.
These fields refer to the most recent transplant. Original coded columns
remain available alongside the harmonized variables.

```{r}
#| label: harmonize-patients
patients_harmonized <- harmonize_patients(patients)
```

## Save harmonized data

```{r}
#| label: save-patients
dir.create("../data/preprocessed", recursive = TRUE, showWarnings = FALSE)

saveRDS(
  patients_harmonized,
  "../data/preprocessed/patients_characteristics_harmonized.rds"
)
write_csv(
  patients_harmonized,
  "../data/preprocessed/patients_characteristics_harmonized.csv",
  na = ""
)
```

```{r}
#| label: summary
patients_harmonized |>
  summarise(
    rows = n(),
    participants = n_distinct(record_id),
    missing_age = sum(is.na(age)),
    transplant_recipients = sum(transplant_recipient == "Yes", na.rm = TRUE)
  ) |>
  knitr::kable()
```

## Run preprocessing

From the project root:

```bash
quarto render R/00_data_preprocess.qmd
```

To use the function directly from R at the project root:

```{r}
#| eval: false
source("R/REVIVE_harmonization.R")
patients <- readr::read_csv(
  "data/Patients_characterisitics.csv",
  col_types = readr::cols(.default = readr::col_character())
)
patients_harmonized <- harmonize_patients(patients)
```

This step prepares patient characteristics only. Infection and vaccination
histories require a separate step to produce the dashboard's
`data/preprocessed/longitudinal_events.json`.
