---
title: "REVIVE: longitudinal data preparation"
format:
  html:
    toc: true
execute:
  echo: true
  cache: false
---

## Purpose

Read the infection and vaccination CSV exports in `data/`, apply the REDCap
dictionary labels, and consolidate reports by **`record_id`** into
`data/preprocessed/longitudinal_events.json`. Source CSVs are never modified.

Each participant has an array of infections and an array of vaccinations.
This preserves multiple events over time without a many-to-many join.
The application JSON also includes `Sex`, `Age`, `Race`, `Immunosuppression`,
and `SOTRs`, coalesced across rows with the same ID. Participants without
reported infections or vaccinations have empty event arrays (`[]`).

## Setup

Run this document from the project root or from `R/`. It requires the
`readr` and `jsonlite` R packages; rendering also requires Quarto and `knitr`.
Reusable transformation functions live in `R/helpers_preprocess.R` and
`R/helpers_participants.R`.

```{r}
#| label: setup
project_root <- if (file.exists("R/helpers_preprocess.R")) {
  normalizePath(".")
} else if (file.exists("../R/helpers_preprocess.R")) {
  normalizePath("..")
} else {
  stop("Run this document from the REVIVE project root or its R directory.")
}
source(file.path(project_root, "R", "helpers_preprocess.R"))
source(file.path(project_root, "R", "helpers_participants.R"))
data_dir <- file.path(project_root, "data")
```

## Dictionary and retained fields

The mappings below are transcribed from
[`ref/RespiratoryViralInfectionVacci.pdf`](../ref/RespiratoryViralInfectionVacci.pdf),
printed page **120** for infections and page **79** for vaccinations.
All six slots use the same mapping within their event category. Labels
come from the field values, not the export filenames: a file named
“FluVaccines” can also contain COVID-19 vaccinations.

```{r}
#| label: dictionary
dictionary <- revive_event_dictionary()
data.frame(
  code = names(dictionary$infection),
  infection = unname(dictionary$infection),
  vaccine_type = unname(dictionary$vaccination)
) |> knitr::kable()
```

| Source | ID | Type fields | Date fields |
|:--|:--|:--|:--|
| Infections | `record_id` | `new_inf1` through `new_inf6` | `new_inf1_dt` through `new_inf6_dt` |
| Vaccinations | `record_id` | `new_vax_recvd`, `new2_vax_recvd` through `new6_vax_recvd` | `new_vax_dt`, `new2_vax_dt` through `new6_vax_dt` |

Slots are discovered from column names, so additional numbered slots follow
the same rules. Completely blank type/date pairs are ignored. Partially
filled reports are retained and flagged. Unknown codes become a `null`
label; missing or invalid dates become `null`. The audit RDS preserves
the original codes/dates and identifies the problem.

IDs remain strings, including leading zeros. Dates are validated and stored
as `YYYY-MM-DD`. Source dates may use `YYYY-MM-DD`, `MM/DD/YYYY`, or
`MM-DD-YYYY`, consistent with the dictionary's month/day/year fields.

## Prepare and validate

The block below selects participant attributes from
`RespiratoryViralInfe_DATA_2026-09-21_1151.csv`, coalesces nonmissing values
within each `record_id`, applies dictionary labels, and joins them to the
event histories by ID. Conflicting nonmissing values stop the run so that
a participant's attributes are never chosen arbitrarily.

| Source fields | JSON variable | Representation |
|:--|:--|:--|
| `sex` | `Sex` | Dictionary label (page 19) |
| `age_calc_consent` | `Age` | Numeric age at consent (page 17) |
| `race___*` | `Race` | Array of selected race labels (page 19) |
| `is_meds_elig___*` | `Immunosuppression` | Array of selected medication labels (page 23) |
| `prior_tx_organ___*` | `SOTRs` | Array of selected **prior** transplant organs |

Checkbox `1` means selected and `0` means unselected. An entirely missing
checkbox group becomes `null`; an observed group with no selections becomes
`[]`. Missing scalar attributes become `null`. The medication choice
`is_meds_elig___0 = 1` explicitly produces `["None"]`.

The PDF embeds `prior_tx_organ` without listing its options on pages 21–22.
For `SOTRs`, the code assumes the organ choices use the same coding as
`organ_tx_28d_type` on page 135; the prior-transplant branching rules confirm
codes 1 (Kidney) and 2 (Liver). This assumption is documented in the helper's
dictionary. `SOTRs` describes prior transplants only, not the most recent
transplant or overall transplant-recipient status.

```{r}
#| label: prepare-data-and-participants
participant_data <- read_revive_participants(file.path(
  data_dir, "RespiratoryViralInfe_DATA_2026-09-21_1151.csv"
))
events <- preprocess_revive_data(
  data_dir,
  output_dir = file.path(data_dir, "preprocessed"),
  participant_data = participant_data
)
participants <- jsonlite::read_json(
  file.path(data_dir, "preprocessed", "longitudinal_events.json"),
  simplifyVector = FALSE
)
```

This is a full join: every ID from either the participant export or the
event reports is retained. The full export also contains event fields, so
the existing event importer includes those reports and removes identical
copies from overlapping exports.

The helper also writes `data/preprocessed/longitudinal_events.rds`: a long
audit table with REDCap event/repeat identifiers, original codes/dates,
source file/row/slot, and quality flags. These fields support review and
are kept out of the smaller application JSON.

```{r}
#| label: summary
# Show aggregate counts only in the rendered report.
data.frame(
  participants = length(participants),
  participants_with_events = length(unique(events$record_id)),
  infection_reports = sum(events$event_type == "infection"),
  vaccination_reports = sum(events$event_type == "vaccination"),
  reports_with_quality_issues = sum(events$quality_issue != ""),
  possible_duplicate_reports = sum(events$possible_duplicate)
) |> knitr::kable()
```

Reports sharing ID, category, type, and date are flagged as possible
duplicates, but distinct REDCap reports are preserved. Resolve these in
the source data before interpreting event counts as unique clinical events.

The JSON uses this structure (illustrative data only):

```json
[
  {
    "record_id": "001",
    "Sex": "Female",
    "Age": 50,
    "Race": ["Asian"],
    "Immunosuppression": ["Tacrolimus (also known as Prograf, Envarsus XR, FK506)"],
    "SOTRs": ["Kidney"],
    "infections": [
      {"infection": "COVID-19 infection", "infection_date": "2026-01-10"},
      {"infection": "RSV infection", "infection_date": "2026-03-12"}
    ],
    "vaccinations": [
      {"vaccine_type": "Flu vaccine", "vaccine_date": "2025-10-01"}
    ]
  }
]
```

## Future updates and Shiny use

1. Add new raw CSV exports under `data/` (subdirectories are supported).
   **Keep the complete source history:** each run rebuilds the output from
   the available CSVs rather than appending to the previous JSON.
2. Export `record_id`, `redcap_event_name`, and matching type/date columns.
   Include `redcap_repeat_instrument` and `redcap_repeat_instance` when
   using repeating REDCap instruments. Each ID/event/repeat/category/slot
   combination identifies one report.
3. Rerun this document. Identical copies of the same report in overlapping
   exports count once. New months, repeat instances, and slots are retained.
   Conflicting versions of the same report stop the run without replacing
   outputs; reconcile outdated exports before rerunning. If the dictionary
   changes, update `revive_event_dictionary()` against the new dictionary.
4. Update the participant CSV filename in the preparation block when using
   a newer full export. Review aggregate quality counts and the audit RDS
   before app analysis.

Render from the project root:

```bash
quarto render R/00_data_preprocess.qmd
```

To generate the same datasets without rendering:

```r
source("R/helpers_preprocess.R")
source("R/helpers_participants.R")
participant_data <- read_revive_participants(
  "data/RespiratoryViralInfe_DATA_2026-09-21_1151.csv"
)
events <- preprocess_revive_data(
  "data", output_dir = "data/preprocessed", participant_data = participant_data
)
```

Read the prepared JSON once at the appropriate app initialization point,
then pass the data to modules:

```{r}
#| label: shiny-loading-example
#| eval: false
participants <- jsonlite::read_json(
  "data/preprocessed/longitudinal_events.json",
  simplifyVector = FALSE
)
```
